Guidance for clinicians in constructing ordinal outcomes and choosing ordinal effect measures for clinical trials

### 2025-05-15 paper.tex
Work before meeting presentation

\section{Desiderata for outcomes and effect measures}

@@

\section{Constructing and combining ordinal outcomes}

\subsection{Fine-graining using tie-breaking and overrides for efficiency}

@@

Two basic operations:
\begin{enumerate}
  \item Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic order)
  \item Tie-breaking by $B$
    \emph{within specific levels of $A$}. This is equivalent to complete
    tie-breaking followed by coarsening.
  \item A special case of this is
    \emph{overriding} of $B$ by $A$ when $A$ is binary.
\end{enumerate}

Examples, using "$/$" to denote overriding and "$\times$" to denote complete
tiebreaking (note that this is an asymmetric operation):
\begin{itemize}
  \item
    ``Angina Symptom Score'' outcome from
    \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
    \ref{fig:rajkumar_et_al_outcome}, is given by \[ \text{death} / \text{ACS} /
      \text{unacceptable angina} / (\text{antianginal meds} \times
        \text{\# angina
    episodes}) \]
  \item Outcome from PASSPORT, depicted in Figure
    \ref{fig:passport_outcome}, is given by \[ \text{death} / \text{SAE} /
      ((\text{max pSOFA} > 1) \times (\text{if max pSOFA > 1: max
            pSOFA else: days
    hospitalized})) \]
\end{itemize}

@@

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{``Angina Symptom Score'' ordinal outocome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) complete tie-breaking (of units of antianginal medication by
    episodes of angina) and (2) overriding (of angina episodes and
      angina medication by adverse clinical events; of all other outcomes
  by death; etc.)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned PASSPORT adaptive
    platform trial. This illustrates (1) overriding (of hospital
    outcomes by pSOFA outcomes, etc.) but also (2) using tie-breaking
    to augment a ``state'' outcome (status at day 28) with
    longitudinal (maximum pSOFA while hospitalized) and time-to-event
    (days hospitalized)
  information}
  \label{fig:passport_outcome}
\end{figure}

@@

\subsubsection{Tie breaking can augment a state outcome with longitudinal
information}

@@

\subsubsection{Overriding can augment a continuous outcome}

For example overriding a continuous outcome by death. It would be
good to have an example of this.

\subsubsection{Precision and power depend mostly on the probability of the
largest category}

We have simulation results and some math to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See \ref{fig:p_max}
for an illustration.

\subsection{Coarsening and incorporating intercurrent events for validity}

@@

\subsubsection{Coarsening for simplicity}

@@

\subsubsection{Incorporating intercurrent events using tie-breaking
or overriding}

@@

@@

What they have in common: the effect measure is a byproduct of fitting a model
to the data.

@@

@@

@@

Rather they are all be based on the ``\emph{pairs construction}'',
where we imagine independently sampling patients from the control and
treatment groups and considering ``wins'' ``ties'' and ``losses''
(corresponding to concordant and discordant pairs and ties in $y$
respectively). They can
also be \emph{estimated from data} using pairs of patients, but this
is not essential and they could also be estimated by fitting a
proportional odds model, for example (though they would not be a
simple function of a coefficient).

@@

\begin{itemize}
  \item Reducing to a binary outcome by picking a specific cutpoint,
    e.g. ``hospitalization or
    worse''. This is effectively using a binary outcome, but with the
    ordinal structure used to gain power if fitting a model.
  \item Reducing to a binary outcome using quantiles of a reference
    group; for example,
    ``worse than the median outcome under standard care''
  \item Ridit analysis
    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
    smithsonReceiverOperatingCharacteristic2023, jansenRiditAnalysisReview1984}
\end{itemize}

@@

These approaches implicitly \textbf{bundle together three different things}:
\begin{enumerate}
  \item Constructing an implicit ordinal outcome through
    complete tie-breaking
  \item Targeting a nonparametric effect measure defined via the
    ``pairs'' device (probabilistic
    index/win ratio/win odds/etc.)
  \item Estimating that effect measure in a nonparametric/flexible
    manner using pairs, often also with devices for accounting for censoring
\end{enumerate}

>>>>

\section{Desiderata for outcomes and effect measures: validity and efficiency}

@@

\section{Combining and coarsening ordinal outcomes}

\subsection{3 ways to combine/fine-grain ordinal outcomes: sequential,
hierarchical, \& parallel composition}

@@

Three broad ways to combine ordinal outcomes:
\begin{enumerate}
  \item \textbf{Sequential composition} Tie-breaking by $B$
    \emph{within either highest or lowest level of $A$}. Also could
    call this ``overriding by $A$''. A special case of this is
    \emph{overriding events} when $A$ is binary.
  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic order) within
    \emph{every} level of $A$.

    This is what win methods do.
  \item \textbf{Parallel composition} This is where levels of $A$ are
    combined with levels of $B$ ``in parallel''; high levels of $A$
    are combined with high levels of $B$ and low levels of $A$ with
    low levels of $B$.

    This is not so precisely determined as the other two. We need to
    decided \emph{how} to aggregate.
\end{enumerate}

Suggested notation:
\begin{itemize}
  \item "$\rightarrow$" and "$\leftarrow$" to denote sequential
    composition/overriding,
  \item "$/$" to denote hierarchical composition/complete tiebreaking
    (note that this is an asymmetric operation), and
  \item "$|$" to denote parallel composition
\end{itemize}

Examples:

\begin{itemize}
  \item \textbf{Sequential composition} Outcome from PASSPORT,
    depicted in Figure
    \ref{fig:passport_outcome}, is given by
    \[ \text{day 28 status} \leftarrow \text{max pSOFA} \leftarrow \text{days
      hospitalized}
    \]
  \item
    \textbf{Hierarchical composition} ``Angina Symptom Score'' outcome from
    \citet{rajkumarPlaceboControlledTrialPercutaneous2023}, depicted in Figure
    \ref{fig:rajkumar_et_al_outcome}, is given by \[
      \text{adverse clinical event} \leftarrow ((\text{units of
    antianginal medication}) / (\text{episodes of angina})) \]

  \item \textbf{Parallel composition} BANDICOOT outcome
    \citep{walkerCodesigningNovelOrdinal2025}, which is
    \[
      (\text{death, graft loss, or relapse}) \leftarrow (\text{viraemia} \;|\;
      \text{CD4+ count} \;|\; \text{organ support})
    \]
\end{itemize}

@@

\subsubsection{Another way: scores}

Treat each input factor as
numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
has examples?

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned PASSPORT adaptive
    platform trial. This illustrates sequential composition/overriding:
    of days hospitalized by maximum pSOFA while hospitalized, and of
  both of these by day 28 status.}
  \label{fig:passport_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{``Angina Symptom Score'' ordinal outocome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) complete tie-breaking (of units of antianginal medication by
    episodes of angina) and (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

@@

\subsection{Reasons to combine ordinal outcomes}

\subsubsection{Incorporating intercurrent events for validity}

For example overriding a numeric outcome by death.
Example: REMAP-CAP: organ-free survival, overridden by death
(Figure \ref{fig:remap_cap_override}).

\begin{figure}
  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
  \caption{Ordinal outcome used in the REMAP-CAP adaptive
    platform trial
    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
    Illustrates overriding a numeric ordinal outcome (organ
    support-free days) by a binary ordinal outcome (death; this is
    ``sequential composition''), which
  would also be an intercurrent event.}
  \label{fig:remap_cap_override}
\end{figure}

\subsubsection{Increasing precision by reducing the probability of the
largest category}

Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
the win probability, and therefore at least approximately to win
difference/win ratio/win odds.

We have simulation results and some math to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
\ref{fig:p_max} for an illustration.

\subsubsection{Augmenting a state outcome with longitudinal
information}

@@

\subsection{Reasons to coarsen ordinal outcomes}

@@

\subsubsection{Coarsening for ease of communication}

@@

@@

\begin{figure}
  \includegraphics[width=7in]{illustrating_common_or.pdf}
  \caption{The common odds ratio, a model-based effect measure, comes
    from the best-fitting parallel curves to the logit transformed
    empirical cumulative distribution. \textbf{Ideally would illustrate
      using toy bar charts as above, with only 2 groups? Although it's a
  strength that model-based measures can compare multiple groups.}}
  \label{fig:illustrating_or}
\end{figure}

@@

What they have in common: the effect measure is a byproduct of fitting a model
to the data. See Figure \ref{fig:illustrating_or} for an illustration
for the common odds ratio.

@@

In Figure \ref{fig:illustrating_or} the pattern assumed to hold is
that the logit-transformed eCDFs in each group are parallel.

@@

\begin{figure}
  \includegraphics[width=7.5in]{effect_measures_pp_plot.pdf}
  \caption{Illustrating independent pairwise comparison-based effect
    measures using a PP-plot or ROC curve. Adapted from
  \citet{smithsonReceiverOperatingCharacteristic2023}.}
  \label{fig:illustrating_pairwise_comparisons}
\end{figure}

@@

Rather they are all be based on the ``\emph{pairs construction}'',
where we imagine independently sampling patients from the control and
treatment groups and considering ``wins'' ``ties'' and ``losses''
(corresponding to concordant and discordant pairs and ties in $y$
respectively). See Figure \ref{fig:illustrating_pairwise_comparisons}
for an illustration.

They can
also be \emph{estimated from data} using pairs of patients, but this
is not essential and they could also be estimated by fitting a
proportional odds model, for example (though they would not be a
simple function of a coefficient).

@@

\begin{itemize}
  \item Dichotomizing by picking a specific cutpoint,
    e.g. ``hospitalization or
    worse''. This is effectively using a binary outcome, but with the
    ordinal structure used to gain power if fitting a model.
  \item Dichotomizing using quantiles of a reference
    group; for example,
    ``worse than the median outcome under standard care''
  \item Simultaneously comparing all quantiles to those of a
    reference group: ridit analysis
    \citep{brossHowUseRidit1958,agrestiAnalysisOrdinalCategorical2010,
      smithsonReceiverOperatingCharacteristic2023,
    jansenRiditAnalysisReview1984}
\end{itemize}

@@

These approaches implicitly \textbf{bundle together three different things}:
\begin{enumerate}
  \item Constructing an implicit ordinal outcome through
    hierarchical composition (complete tie-breaking/prioritized outcomes)
  \item Targeting a nonparametric effect measure defined via the
    ``pairs'' device (win probability/win ratio/win odds/etc.)
  \item Estimating that effect measure in a nonparametric/flexible
    manner using pairs, often also with devices for accounting for censoring
\end{enumerate}
